### 2024-04-07 Chapters/Chapter2.tex

Veðurstofan provided recordings from 412 weather stations all around Iceland. The location of these weather stations can be seen in Figure \ref{fig:aws_map}. These automatic weather stations (AWS) are at a height of 10 meters above ground level. The information that is provided by these AWS is presented in two different type of documents, hourly and 10 minute documents. The hourly documents are summations of the 10 minute documents, with the exception that any errors (so called nails) still in the 10 minute documents should have been removed from the hourly documents. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another document), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction. These features for each of the stations go back at least two decades, but do not have the same start date. This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

\begin{figure}[h]
    \caption{Locations of Automatic Weather Stations in Iceland}
    \label{fig:aws_map}
    \includegraphics[scale = 0.75]{Figures/weather_stations.png}
\end{figure}

@@

After using this method to request terabytes of data, it was discovered that it is apparently possible to query a specific area. This decreased the size of each file created after a request from around 50 MB to around 2 MB. Downloading and processing this data would take a lot less time. The bottleneck for retrieving the data is the request time. That is the time the request is queued and running (server preparing the data) before it begins downloading. This could range from almost immediate to 30 minutes. This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during the winter and spring of 2023/2024, which increased the wait time and sometimes resulted in queries not being responded to. This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project. Fortunately these problems were only particularly time consuming during mid winter.

>>>>

Veðurstofan provided 10 minute recordings from \nStationsMin weather stations all around Iceland. Of these \nStationsMin stations, \nVedurMin were from Veðurstofa Íslands and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{Vegagerðin} and placed at 7.5 meters above ground. The location of these weather stations can be seen in Figure \ref{fig:aws_map}. The information that is provided by these Automatic Weather Stations (AWS) is presented in two different type of documents, hourly and 10 minute documents. The hourly documents are summations of the 10 minute documents, with the exception that any errors (so called nails) still in the 10 minute documents should have been removed from the hourly documents. Nails, are sharp increases from the rest of the data and are thought to be measurements errors by experts and are discarded by hand. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another document), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction. These features for each of the stations go back at least two decades, but do not have the same start date. This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

\begin{figure}
    \caption{Locations of Automatic Weather Stations in Iceland}
    \label{fig:aws_map}
    \includegraphics[scale = 0.75]{Figures/weather_stations.png}
\end{figure}

These 10 minute measurements were filtered to only include measurements where the average wind speed was at or above \averageWindSpeedLimit m/s. They were also filtered in such a way that only the highest value in any given 48 hour span is considered. This can be seen in Code Listing (\ref{code:filter48hr}). Filtering by, for each station, the highest wind speed and removing all observations that occurred within a 48 hour period centered on the observation (including the highest wind) and doing this iteratively until reaching an empty dataframe for a given station.

\begin{lstlisting}[language = Python, caption = {Filter points over 48 hour interval}, label = code:filter48hr]
    for station in tqdm(stations, total = len(stations)):
        subset_df = vedur_df[station == vedur_df.stod]
        subset_df = subset_df.reset_index(drop = True)

        while not subset_df.empty:
            idx = subset_df.f.idxmax()
            time_of_max = subset_df.iloc[idx].timi

            filtered_data.append(subset_df.iloc[idx])

            subset_df = subset_df[abs(subset_df.timi - time_of_max)
                >= pd.Timedelta(threshold)]

            subset_df = subset_df.reset_index(drop = True)  
\end{lstlisting}

When these two filterings have been done, the end result is \nStationsCombined stations that meet these criteria. Of these, \nVedurMin belong to Veðurstofa Íslands and \nVGMin belong to Vegagerðin. The number of observations after, these filterings, by year can be seen in Figure (\ref{fig:occurrences_by_year}). The first observations occurring in \startDateVedur and the latest in 2023.

\begin{figure}
    \caption{Measurements that fit filtering by year}
    \label{fig:occurrences_by_year}
    \begin{center}
        \includegraphics[scale = 0.75]{Figures/occurrences_by_year.png}
    \end{center}
\end{figure}

Looking at the number of measurements that meet these criteria for a given station, Figure (\ref{fig:occurrences_by_station_20}) shows only one station has over 1000 measurements that fit the criterias.

\begin{figure}
    \caption{Number of stations that have measurements in a range}
    \label{fig:occurrences_by_station_20}
    \begin{center}
        \includegraphics[scale = 0.75]{Figures/station_occurrences_20.png}        
    \end{center}
\end{figure}

@@

After using this method to request terabytes of data, it was discovered that it is apparently possible to query a specific area. This decreased the size of each file created after a request from around 50 MB to around 2 MB. Downloading and processing this data would take a lot less time. The bottleneck for retrieving the data is the request time. That is the time the request is queued and running (server preparing the data) before it begins downloading. This could range from almost immediate to 30 minutes. This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during \href{https://forum.ecmwf.int/t/a-new-cds-soon-to-be-launched-expect-some-disruptions/1607}{the winter and spring of 2023/2024}, which increased the wait time and sometimes resulted in queries not being responded to. This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project. Fortunately these problems were only particularly time consuming during mid winter.

### 2024-04-07 Chapters/Chapter3.tex

This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models. When working with hundred of thousands of rows, the efficiency of the code is very important. Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible. The three data sources were all in a different format. Measurement data from Veðurstofa was in text files, elevation data was in GeoTiff and reanalysis data from Carra was in a GRIB format. To use the data to train, these three data sources needed to be combined into one file. This was done based on the measurement data from the Veðurstofa. A limit was set on the average wind speed and it was used to select measurement points. Along with the average wind speed having to be above a certain limit, to make sure that the same weather for the same location is not duplicated, only the top wind speed in any given 48 hour period is selected. That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times. The data from Veðurstofan was supplied for 10 minute increments, while Carra data is in 3 hour intervals. This means that to use the Carra data to predict the measured values from Veðurstofa, temporal interpolation would need to be done. Along with the temporal interpolation, note that the Carra data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations. The elevation information was given by a 20 by 20 rectangular grid that covers Iceland. When combining these data sources a selection of interpolation method needs to occur. The interpolation was linear, both temporally and spatially. This might influence the results but was not considered in this study.

@@

The procedure of combining these sources was as follows and can be seen in Figure (\ref{fig:data_preprocessing_flow_chart}). The measured data from the AWS is filtered by using a limit on the average wind speed. The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}). Even so being able to predict the gust factor is more important for higher average wind speed as there are the highest wind gusts. After this stripped dataset over every AWS has been created it is used to query the Carra data by using their API. The Carra API needs to be queried for given hours, days, months, years and a given area. That is, if queried for a given hour, it returns that hour for every day that is queried. Similarly if queried for a given day, it returns that day for every month. In light of these restraints, it was decided to query month by month. Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00 PM) and then take these values and interpolate for the 10 minute values. After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe. After this the downloaded data is discarded and the next month is queried. This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).

The elevation data comes in a GeoTif file that covers Iceland. It is a rectangular grid of resolution 20 meters. For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved. For each point retrieved interpolation needs to be done. This is done in a similar manner to the interpolation of the Carra data. Weights are assigned to each of the four points bounding the point of interest. These weights are then used to calculate the weighted average that represents the interpolated value. This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.

@@

Once data has been retrieved for all three sources and processed, including interpolating values, it needs to be made ready to use by the model, for both training, validation and test. Starting with a dataframe that contains measured information from AWS. This includes the average wind, the wind gust, wind direction along with the station number and coordinates. When selecting the Carra data certain height levels are chosen. That is at which heights above ground the reanalysis parameters are requested. These present as separate lines in the Carra dataframe. That is information for one observation is represented in as many lines as height levels requested in the reanalysis data. These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station. When this is done it is possible to combine the AWS Veðurstofu data and Carra reanalysis data on the location and time columns. The last data source is the elevation. A couple of different sections of land around the weather stations have been looked at. A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point. In any case the points, that represent these sections, were selected like as shown in code Listing (\ref{code:sectorElevation}). That is a range of angles are defined based on the wind direction d at some distance from the given point. This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.

@@

The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from Carra, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data. An example of what the data looks like can be seen in Table (\ref{table:trainDataExample}).

>>>>

This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models. When working with hundred of thousands of rows, the efficiency of the code is very important. Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible. The three data sources were all in a different format. Measurement data from Veðurstofa was in text files, elevation data was in GeoTiff and reanalysis data from CARRA was in a GRIB format. To use the data to train, these three data sources needed to be combined into one file. This was done based on the measurement data from the Veðurstofa. A limit was set on the average wind speed and it was used to select measurement points. Along with the average wind speed having to be above a certain limit, to make sure that the same weather for the same location is not duplicated, only the top wind speed in any given 48 hour period is selected. That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times. The data from Veðurstofan was supplied for 10 minute increments, while CARRA data is in 3 hour intervals. This means that to use the CARRA data to predict the measured values from Veðurstofa, temporal interpolation would need to be done. Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations. The elevation information was given by a 20 by 20 rectangular grid that covers Iceland. When combining these data sources a selection of interpolation method needs to occur. The interpolation was linear, both temporally and spatially. This might influence the results but was not considered in this study.

@@

The procedure of combining these sources was as follows and can be seen in Figure (\ref{fig:data_preprocessing_flow_chart}). The measured data from the AWS is filtered by using a limit on the average wind speed. The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}). Even so being able to predict the gust factor is more important for higher average wind speed as there are the highest wind gusts. After this stripped dataset over every AWS has been created it is used to query the CARRA data by using their API. The CARRA API needs to be queried for given hours, days, months, years and a given area. That is, if queried for a given hour, it returns that hour for every day that is queried. Similarly if queried for a given day, it returns that day for every month. In light of these restraints, it was decided to query month by month. Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00 PM) and then take these values and interpolate for the 10 minute values. After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe. After this the downloaded data is discarded and the next month is queried. This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).

Once CARRA data has been merged with AWS data, using station and time columns, then this combined file needs to checked for nails. This is done by using the hourly data (which is supposedly error free). As the data has been filtered in for the highest average wind speed in a 48 hour interval, the hourly data can be used to find nails. The hourly data is combined with the merged AWS and 10 min data. Then filtering is applied on the average wind. If the average wind speed differs the rows are dropped. This can be used to look at which stations have high percentage error. This can be seen in Figure (\ref{fig:error_count}).

\begin{figure}
    \caption{The number of stations whose number of nails are in a certain percentage range of the overall}
    \label{fig:error_count}
    \includegraphics[scale=0.75]{Figures/error_count.png}
\end{figure}

Figure (\ref{fig:error_count}) shows that most stations have nails in less that 10\% measurements. One has nails in 100\% of measurements. This is only because the number of measurements that matched previous criteria was 1.

The elevation data comes in a GeoTif file that covers Iceland. It is a rectangular grid of resolution 20 meters. For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved. For each point retrieved interpolation needs to be done. This is done in a similar manner to the interpolation of the CARRA data. Weights are assigned to each of the four points bounding the point of interest. These weights are then used to calculate the weighted average that represents the interpolated value. This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.

@@

Once data has been retrieved for all three sources and processed, including interpolating values, it needs to be made ready to use by the model, for both training, validation and test. Starting with a dataframe that contains measured information from AWS. This includes the average wind, the wind gust, wind direction along with the station number and coordinates. When selecting the CARRA data certain height levels are chosen. That is at which heights above ground the reanalysis parameters are requested. These present as separate lines in the CARRA dataframe. That is information for one observation is represented in as many lines as height levels requested in the reanalysis data. These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station. When this is done it is possible to combine the AWS Veðurstofu data and CARRA reanalysis data on the location and time columns. The last data source is the elevation. A couple of different sections of land around the weather stations have been looked at. A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point. In any case the points, that represent these sections, were selected like as shown in Code Listing (\ref{code:sectorElevation}). That is a range of angles are defined based on the wind direction d at some distance from the given point. This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.

@@

The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from CARRA, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data. An example of what the data looks like can be seen in Table (\ref{table:trainDataExample}).
